A Tutorial on Tailored Simulation-Based Sample-Size Planning for Experimental Designs with Generalized Linear Mixed Models

### 2024-11-06 _manuscript/_tex/manuscript.tex
update citation metadata; small fixes

\title{A Tutorial on Tailored Simulation-Based Sample Size Planning for
Experimental Designs with Generalized Linear Mixed Models}

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, left=2mm, title=\textcolor{quarto-callout-important-color}{\faExclamation}\hspace{0.5em}{Important}, colframe=quarto-callout-important-color-frame, breakable, rightrule=.15mm, coltitle=black, titlerule=0mm, arc=.35mm, toprule=.15mm, bottomtitle=1mm, colbacktitle=quarto-callout-important-color!10!white, opacitybacktitle=0.6, bottomrule=.15mm, toptitle=1mm, opacityback=0, colback=white]

@@

Fourth, all frameworks for sample size planning require the user to make
assumptions about the expected effect size. Assuming the true effect is
of this size (or greater), one can compute the (minimum) power of a HT
or the (maximum) expected width of a CI. Existing software packages for
sample size planning for GLMMs usually require to provide the assumed
effect in the unit of some standardized measure of effect size. When the
researcher has access to similar studies or pilot data, providing such
standardized effect sizes is feasible. However, choosing effect sizes
based on small pilot studies is generally not recommended, as those
estimates can be heavily biased
(\citeproc{ref-albersWhenPowerAnalyses2018a}{Albers \& Lakens, 2018};
\citeproc{ref-lakensSampleSizeJustification2022}{Lakens, 2022a}).
Providing an informed standardized effect size can be an almost
impossible challenge when no prior studies of pilot data are available.
This problem is further exacerbated by the fact that GLMMs are so
flexible that general heuristics of what should be considered a small
effect do not exist or are difficult to defend. In the absence of prior
evidence, using domain knowledge to construct a tailored data simulation
is considered the best solution to determine plausible effect sizes (see
scenario 3 in \citeproc{ref-kumleEstimatingPowerGeneralized2021}{Kumle
et al., 2021}). It would be possible to use these tailored simulations
to compute standardized effect sizes that could then be inserted in
existing software packages for sample size planning. However, we would
argue that when tailored data simulations are necessary to determine
effect sizes anyway, performing the whole sample size planning in a
customized way is preferred over using the existing software packages.

>>>>

\title{A Tutorial on Tailored Simulation-Based Sample-Size Planning for
Experimental Designs with Generalized Linear Mixed Models}

@@

\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, bottomtitle=1mm, arc=.35mm, toptitle=1mm, left=2mm, opacityback=0, colbacktitle=quarto-callout-important-color!10!white, colback=white, opacitybacktitle=0.6, rightrule=.15mm, breakable, title=\textcolor{quarto-callout-important-color}{\faExclamation}\hspace{0.5em}{Important}, colframe=quarto-callout-important-color-frame, titlerule=0mm, bottomrule=.15mm, coltitle=black, leftrule=.75mm]

@@

Fourth, all frameworks for sample size planning require the user to make
assumptions about the expected effect size. Assuming the true effect is
of this size (or greater), one can compute the (minimum) power of a HT
or the (maximum) expected width of a CI. Existing software packages for
sample size planning for GLMMs usually require to provide the assumed
effect in the unit of some standardized measure of effect size. When the
researcher has access to similar studies or pilot data, providing such
standardized effect sizes is feasible. However, choosing effect sizes
based on small pilot studies is generally not recommended, as those
estimates can be heavily biased
(\citeproc{ref-albersWhenPowerAnalyses2018a}{Albers \& Lakens, 2018};
\citeproc{ref-lakensSampleSizeJustification2022}{Lakens, 2022a}).
Providing an informed standardized effect size can be an almost
impossible challenge when no prior studies or pilot data are available.
This problem is further exacerbated by the fact that GLMMs are so
flexible that general heuristics of what should be considered a small
effect do not exist or are difficult to defend. In the absence of prior
evidence, using domain knowledge to construct a tailored data simulation
is considered the best solution to determine plausible effect sizes (see
scenario 3 in \citeproc{ref-kumleEstimatingPowerGeneralized2021}{Kumle
et al., 2021}). It would be possible to use these tailored simulations
to compute standardized effect sizes that could then be inserted in
existing software packages for sample size planning. However, we would
argue that when tailored data simulations are necessary to determine
effect sizes anyway, performing the whole sample size planning in a
customized way is preferred over using the existing software packages.
